---
output: github_document
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = T)
```

# grpsel

[![R-CMD-check](https://github.com/ryan-thompson/grpsel/workflows/R-CMD-check/badge.svg)](https://github.com/ryan-thompson/grpsel/actions)
[![test-coverage](https://github.com/ryan-thompson/grpsel/workflows/test-coverage/badge.svg)](https://github.com/ryan-thompson/grpsel/actions)

## Overview

An R package for sparse regression modelling with grouped predictors (including overlapping groups). `grpsel` makes use of the group subset selection penalty, which usually leads to excellent selection and prediction. Optionally, the group subset penalty can be combined with a group lasso or ridge penalty for added shrinkage. Linear and logistic regression are currently supported.

## Installation

To install the latest stable version from CRAN, run the following code:
``` {r, eval = F}
install.packages('grpsel')
```

To install the latest development version from GitHub, run the following code:
``` {r, eval = F}
devtools::install_github('ryan-thompson/grpsel')
```

## Usage

The `grpsel()` function fits a group subset regression model for a sequence of tuning parameters. The `cv.grpsel()` function provides a convenient way to automatically cross-validate these parameters.
```{r, example} 
library(grpsel)

# Generate some grouped data
set.seed(123)
n <- 100 # Number of observations
p <- 10 # Number of predictors
g <- 5 # Number of groups
group <- rep(1:g, each = p / g) # Group structure
beta <- numeric(p)
beta[which(group %in% 1:2)] <- 1 # First two groups are nonzero
x <- matrix(rnorm(n * p), n, p)
y <- x %*% beta + rnorm(n)

# Fit the group subset selection regularisation path
fit <- grpsel(x, y, group)
coef(fit, lambda = 0.05)

# Cross-validate the group subset selection regularisation path
fit <- cv.grpsel(x, y, group)
coef(fit)
```

See the package [vignette](https://cran.r-project.org/web/packages/grpsel/vignettes/vignette.html) or [reference manual](https://cran.r-project.org/web/packages/grpsel/grpsel.pdf) for more help getting started.

